---
title: "Common_ggplot_geoms"
author: "HDS"
date: "2023-03-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Producing different types of plots with ggplot and geoms

Get the neighborhood data for Buffalo

# Data Set- Neighborhoods in Buffalo

```{R}
infile="C:\\Users\\hdavi\\Dropbox\\Data_Analytics\\DAT511_Data_Cleaning\\Lectures_Spring_2023\\Examples\\Neighborhood_Metrics.csv"
ndata=read.csv(infile,stringsAsFactors=TRUE)
head(ndata)
```

Load ggplot and set up to use the WSJ theme colors

```{R}
require(ggplot2)
require(ggthemes)
theme_set(theme_wsj())
```

#Plots to look at the behavior of a single variable

Many times, we will want to focus on a single variable to understand the distribution of that single variable,  or how it changes relative to a factor

https://r-graph-gallery.com/index.html

# Boxplot

Median income by Community

Boxplots show the median value, the 25% and 75% values and the 95% range, plus any outliers

```{R}
ggplot(ndata,aes(x=Community,y=Median.Income))+geom_boxplot()
```

#Violin Plot

This is a variant on the Boxplot, in which the width of the outline shows the number of values a the vertical height

```{R}
ggplot(ndata,aes(x=Community,y=Total.Persons))+geom_violin()+ggtitle("Total Persons per Neighborhood by Community")
```

#Density Plot

This is a histogram normalized to 1, so the area under the curve is equal to 1

The curve height is an estimate of the probability of taking on the value along the x-axis

```{R}
ggplot(ndata,aes(x=Median.Income))+geom_density(fill="blue")+ggtitle("Median Income, Communities")
```
# Basic Histogram

```{r}
ggplot(ndata,aes(x=Median.Income))+geom_histogram(binwidth=5000)+ggtitle("Histogram of Median Income")
```

#Comparing two variables, Looking at Correlation or Covariance

Working with two variables at a time is called bivariate statistics

More than two is multivariate statistics.

Bivariate

These plots are meant to reveal the relationship between two variables

#Scatter plots

We've seen these before....

The WSJ theme cut off the x and labels, for reason I don't understand, so I changed theme here

```{R}
ggplot(ndata,aes(x=Percent.Age.65..,y=Poverty.Rate))+geom_point()+xlab("Percent over 65")+ylab("Percent Poverty")+ggtitle("Poverty and Age")+theme_economist_white()

```
# Adding a linear trend to a scatter plot

With the trend and the confidence interval on the trend

The method used here was a linear curve fit,  and the confidence interval showing the level of uncertainty in the model is also show.

Other approaches to estimating the curve are possible

```{R}
ggplot(ndata,aes(x=Percent.Age.65..,y=Poverty.Rate))+geom_point()+xlab("Percent over 65")+ylab("Percent Poverty")+ggtitle("Poverty and Age")+theme_economist_white()+geom_smooth(method=lm,color="red",fill="grey",se=TRUE)
```
# 2D density plot,  shows the number of points within a binned region of the plot


```{R}
ggplot(ndata,aes(x=Percent.Age.65..,y=Poverty.Rate))+geom_hex(bins=8)+xlab("Percent over 65")+ylab("Percent Poverty")+ggtitle("Poverty and Age")+theme_economist_white()
```
This data set has relatively few points,  so this plot doesn't look great

Let's try it with the Iris data set

The color of each hexagram shows the number of points within that region

```{R}
data(iris)
ggplot(iris,aes(x=Petal.Length,y=Petal.Width))+geom_hex(bins=25)+ggtitle("Iris data")+theme_economist_white()+scale_fill_continuous(type='viridis')
```

# Bubble Plots

The Bubble plot is a scatter plot, but with a third variable controlling the size of the symbol-so this shows the interaction of three variables

We simple use a variable to control size (Percent.Living.Alone in this case)

We will also set color to the community

the parameter alpha controls the transparency of the data points

```{R, fig.width=16, figure.height=16}
ggplot(ndata,aes(x=Percent.Age.65..,y=Poverty.Rate, size=Percent.Living.Alone, color=Community))+geom_point(alpha=0.7)+xlab("Percent over 65")+ylab("Percent Poverty")+ggtitle("Poverty and Age")+scale_size(range = c(3, 18), name="% living alone")
#+theme_economist_white()
```









# Heatmaps

Heatmaps are a simple way to visualize the correlation patterns in data, the
extent to which two variables are correlated or change in unison

We want to use continuous variables (not factors) in a heat map

Looks look at these variables

Poverty.Rate
Employment.Rate
Percent.Age.65..
Labor.Force.Participation.Rate
Percent.Living.Alone
Percent.Female.Persons

We need to put them into their own data frame to work with them
I'm also going to include Neighborhood and Community as factors, we may want them later

This is a base R function, rather than a ggplot function

```{R,fig.height=8}
np_vars=c("Neighborhood","Community","Poverty.Rate","Employment.Rate","Percent.Age.65..","Labor.Force.Participation.Rate","Percent.Living.Alone", "Percent.Female.Persons")

npdata=ndata[,np_vars]

# in the heatmap, I want to drop the first two columns, which are factors 

heatmap(cor(npdata[,c(-1,-2)]))
```

```{R}
library(corrgram)

corrgram(npdata,order=TRUE, lower.panel=panel.shade, upper.panel=panel.pts, text.panel=panel.txt)
```

# Barplots

This is a flipped barplot

The flip was doing using coord_flip()

Community was used as the "fill" value in the Aesthetic, something we haven't seen before

```{R,fig.height=16,fig.width=16}
ggplot(ndata,aes(x=Neighborhood,y=Poverty.Rate,fill=Community))+geom_bar(stat='identity')+coord_flip()

```

# Reordering factors by some other variable

It would be nice to plot this by decreasing poverty rate

We can reorder the data to do this

The reordering here is odd, because we need to reorder the factors in ndata.

Each factor is stored as an integer, and ggplot starts with the first factor value (1) and proceeds from there

When the factors are created, each neighborhood name is assigned a code, based on the order the name appears in 
the dataframe.    The categories were ordered when the data was imported, so we need to force R to change the order of the factors

To do this, we can sort the data by Poverty.Rate,   the variable we want to order the graph by

Then we reset the factor names by using the command 

mutate(Neighborhood=factor(Neighborhood,levels=Neighborhood))

which resets Neighborhood as a factor, but with the order controlled by the dataframe sorted by Poverty.Rate

so the neighborhood factor values are now ordered by the variable we want for the plot

This seems a bit convoluted, but it does allow fine level of control of the plot

```{R,fig.height=16,fig.width=16}

require("dplyr")

# reorder and plot


nfdata=ndata%>%arrange(Poverty.Rate)%>%mutate(Neighborhood=factor(Neighborhood,levels=Neighborhood))

ggplot(nfdata,aes(x=Neighborhood,y=Poverty.Rate,fill=Community))+geom_bar(stat='identity')+coord_flip()

```

#Piecharts

ggplot has no direct way to create piecharts, you create a barplot then plot it in circuluar coordinates, as shown 
in the examples below

#A note about piecharts

Piecharts are a very divise topic,  weirdly enough.    Most statisticians absolutely hate them.

The problem is that they represent data as angles (slices of the pie) and humans are poor at estimating angles.

That means that while many parts of the population like Pie charts, they are not reliable ways to represent information.

In some parts of the business world,  leaders demand piecharts.   Not much you can do about it.

Try to use barcharts instead whenever possible

See

https://www.data-to-viz.com/caveat/pie.html

```{R}
# Create Data
tempdata <- data.frame(
  group=LETTERS[1:5],
  value=c(13,7,9,21,2)
)

# Basic piechart
ggplot(tempdata, aes(x="", y=value, fill=group)) +
  geom_bar(stat="identity", width=1) 
```


```{R}
# Basic piechart
ggplot(tempdata, aes(x="", y=value, fill=group)) +
  geom_bar(stat="identity", width=1) +
  coord_polar("y", start=0)
```

#treemap as an alternative to barcharts

In a treemap,  the data is represented as squares,   people do better judging the size of rectangles

```{R}
require(treemap)

treemap(ndata,
            
            # data
            index="Neighborhood",
            vSize="Total.Households",
            type="index",
            
            # Main
             title="Households per Neighborhood",
             palette="Dark2",

            # Borders:
             border.col=c("black"),             
             border.lwds=1,                         
        
            # Labels
            # fontsize.labels=0.5,
            # fontface.labels=1,            
            # bg.labels=c("transparent"),              
             align.labels=c("left", "top"),                                  
             overlap.labels=0.5,
             inflate.labels=FALSE                        # If true, labels are bigger when rectangle is bigger.

            
            )


```

Just for the sake of amusement, here's a pie chart of the households per neighborhood



```{R,fig.height=16,fig.weight=16}
# Basic piechart
ggplot(ndata, aes(x="", y=Poverty.Rate, fill=Neighborhood)) +
  geom_bar(stat="identity", width=1) +
  coord_polar("y", start=0)
```