---
title: "Breaking Up Strings"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Break Apart Strings

Also called parsing

We often will need to break up strings to extract portions of them.

Suppose we had a set of strings for each customer

"Name: Joe Smith"
"Address: 3030 Main St"
"City: Buffalo"
"State: NY"

And we wanted to find the state Joe Lives in.

Here the ":" is used as delimiter

We could first find the line that has the State

```{r}
customer=c("Name: Joe Smith",
"Address: 3030 Main St",
"City: Buffalo",
"State: NY")
```

We will use grep to find the line in customer that has the State information

```{r}
state_loc=grep("State",customer)
state_loc
```
Then we want to split the string apart to get the value of the state

state_loc tells us which line has the state, we can then split it on the delimiter ":"

```{r}
temp=strsplit(customer[state_loc],":")
temp
```
strsplit returns something a bit odd, what is this?

```{r}
class(temp)
```
It is a list,  where each entry is the result of splitting a single string.  We only split one string so there is only one item in the list


```{r}
(temp[[1]])
```
To get at the state, we want the 2nd string here in the first location of the list

```{r}
my_state=temp[[1]][2]
my_state
```
We hae a leading space, we can remove that we can use trimws to remove leading and trailing whitespace

```{r}
require("stringi")
my_state=trimws(my_state)
my_state
```
The other approach to extracting text from a string is use the substr() function that extracts characters by location from the string.  So we need to know the location of the state in the string.

We know the state is in the string customer[state_loc], we need the location within the string

we can use regex to find the location of ": " that starts the state, and use length(customver[state_loc]) as the end location

loc1=regexpr(": ",customer[state_loc]) will tell us the location of of ": ",  the state starts two characters after that

```{r}
require("stringr")
loc1=regexpr(": ",customer[state_loc])
print(loc1)
loc2=str_length(customer[state_loc])
temp=substr(customer[state_loc],loc1[1]+2,loc2)
temp
```
The approach based on using regexpr is a bit more complicated, but strsplit
allows only one simple delimiter,  regexpr allows use of a complex regular expression to
locate the segment of string desired.

If you wanted to extact the number from a string like these

"Bob has 12 new potatoes"

"Sue purchased 35 tomatoes"

Splitting these on space doesn't help much, we need to know where the number is

```{r}

a="Bob has 12 new potatoes"

b="Sue purchased 352 tomatoes"

n_loc=gregexpr('[[:digit:]]+',a)
print(n_loc)

```
Here n_loc is an integer with "attributes"

```{r}
print(n_loc[[1]])
print(attributes(n_loc[[1]]))
```

to extract the number, we need the location of the first digit and the length, 

```{r}

loc1=n_loc[[1]]
loc2=attributes(n_loc[[1]])$match.length + loc1-1
substr(a,loc1,loc2)

```
Will this work for string b?

```{r}
n_loc=gregexpr('[[:digit:]]+',b)
loc1=n_loc[[1]]
loc2=attributes(n_loc[[1]])$match.length + loc1-1
substr(b,loc1,loc2)

```