---
title: "Regular_expressions_part2"
author: "HDS"
date: "10/7/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Regular Expresssions part two

tolower(), toupper()

Humans recognize that the same words with differents in capitalization mean the same thing.  But computers are storing these as different binary codings

before you attempt to look for a match,  convert the data to all upper or lower,  often lower

```{r}
"bob"=="BOB"
"bob"==tolower("BOB")
toupper("bob")=="BOB"
```

Suppose we want to check a list of color values, and the only allowed color are red and blue

We are going to allow r and b as valid abbreviations

The check might look like this

-use a tolower to clean up the values
-look for set membership

```{r}
valid_colors=c('red','blue','r','b')
color_entries=c("Red","Blue","r","R","B","BLUE","bl","b")
tolower(color_entries)%in%valid_colors
```

#strsplit

This function splits a string at the requested delimiter and returns the section before and after the split.
It works well at breaking up sections of text

```{r}
a=strsplit("Name-Johnny","-")
a[[1]][1]
a[[1]][2]
```
```{r}
a=strsplit(c("716-888-2587","514-818-3134","667-811-9212"),"-")
print(a)
matrix(unlist(a),ncol=3,byrow=TRUE)
b=data.frame(matrix(unlist(a),ncol=3,byrow=TRUE))
colnames(b)<-c("Area","Exchange","Last_4")
b
```
#Approximate String matching

The amatch function does approximate string match, it allows some level of error.

It uses one of a number of string matching algorithms to measure how far apart two strings are.  This version is using an osa (Optimal String Alighnment) which is a Levenshtein distance, but also allowing transposition of adjacent characters, so each insertion, deletion, subsitution or swap of adjacent letters is one step

```{r}
require('stringdist')
name_list=c("John Smith","Jon Smith","Johnny Smith","John Smyth","Juan Smith","Johnny Smythe","Jhon Smith","Jhon Smiht")
amatch(name_list,"John Smith",method="soundex",maxDist=1)
```
```{r}
phonetic(name_list)
phonetic(c("Bob Jones","John Smith", "Jane Smythe"))
```





