---
title: "R Notebook"
output: html_notebook
author: "Mike Kozlowski"
---

# Dplyr and sqldf Examples

These are both alternative ways of filtering or searching r dataframes

Dplyr is a unique set of tools

sqldf is SQL for data frames,  so you can use SQL commands on an r dataframe

#Data Pliers

Data Pliers (dplyr) is a R package for manipulating data.

It was largely written by Hadly Wickham,  who was head of the organization that produced
RStudio and GGPlot2

GGplot2 was based on a "grammar of graphics",  which is to say, a form of expressing ideas
about how graphic visualizations of data are stated and formed.

dplyr is based on a "grammar of data manipulation"

There are "native" ways in R to do all the manipulation tasks that are carried out
in dplyr.    The "Native R" approaches are very similar to "Native Python" approaches.  The syntax is a bit different, but the general ideas are the same and they work pretty much the
same way.   The python Pandas package works much like Native R.

I think there are some advantages to dplyr, it does some things very well.   In other cases, I find "Native R" to work better.

Personally,   I do not use dplyr.   It takes time to learn, and it is R only.    The native R and Python approaches work just fine.

But, in some subcultures,  dplyr is very popular, so you need to have some exposure to it.

If you like it, go ahead and use it.

# Tidyverse

Hadley Wickham has had a bunch of ideas about processing data,  a lot of them are very
interesting (as is dplyrs).    They are packaged up into a library called tidyverse, so you can just download them all at once.

See

https://dplyr.tidyverse.org/

# Basic Operations in dplyr
---------------------------------------

mutate() adds new variables that are functions of existing variables
select() picks variables based on their names.
filter() picks cases based on their values.
summarise() reduces multiple values down to a single summary.
arrange() changes the ordering of the rows.

```{R}
require("dplyr")
data(starwars)
head(starwars)
```

This example data set is in the form of a tibble.   A tibble is like a data frame
except that it is stricter in how it handles data, so it sort of a more reliable dataframe

-you cannot change variable names or data types of columns
-it creates errors if you search on a variable that does not exist
-it does not allow recycling of variables

# dplyr filtering

```{R}
starwars %>% filter(species=="Droid")
```

The tibble starwars was pipelined into a filter operation,  that selected for species equal to Droid.    The pipeline symbol is %>% and it passes a tibble through a series of operations
in dplyr

The pipeline operator idea is really a big part of unix and linux,  and you may see it later
in python for some more advanced operations.   You can link togehter a bunch of operations with a pipeline

The native R version of this filter operation,  just for comparison

```{R}
starwars[starwars$species=="Droid",]
```
We can pipeline two filters

```{r}
starwars%>%filter(species=="Droid")%>%filter(height>100)
```

In base R

```{R}
starwars[(starwars$species=="Droid")&(starwars$height>100),]
```

#dplyr select

filters variables by name

This allows some interesting combinations

```{R}
starwars%>%select(name,ends_with("color"))
```
In base R we would do this

```{r}
ctargets=c("name","hair_color","skin_color","eye_color")
starwars[,ctargets]

```

#dplyr mutate

This adds a new, computed column to the data

Note this set of commands creates bmi and then selects a few columns

```{r}
starwars %>% 
  mutate(name, bmi = mass / ((height / 100)  ^ 2)) %>%
  select(name:mass, bmi)
```

We cannot simply add a column to a tibble,  we have use mutate

We can add columns to a dataframe

```{R}
df_starwars<-as.data.frame(starwars)
df_starwars["bmi"]<-df_starwars$mass/(df_starwars$height/100)^2
df_starwars[,c("name","height","mass","bmi"),]

```

#arrange()

sort or order data

This has the ability to use complicated orderings that are hard to do in
base R

This is a simple calculation

```{R}
starwars %>% 
  arrange(desc(mass))
```

```{R}
# create an ordering index based on mass
temp=order(df_starwars$mass,decreasing=TRUE)

df_starwars[temp,]
```

One thing that dplyr can do that is tough in base R is a compound ordering, such
as we see in Excel

Remember this if you need it!

```{R}
starwars %>% 
  arrange(species,desc(mass))
```

# summarise and group_by

Here are two operations working together

This is much like a tapply operation in base R

using group_by to group by a factor

then summarise calculates a couple of variables for us
n() is a count operation within summarise,  mean is a the standard mean function

We can set up a number of different operations within summarise

This would be a bit tough to do in base r.

Tapply can group data, but it only allows one function,  we could call tapply several times
and merge the data

```{R}
starwars %>%
  group_by(species) %>%
  summarise(
    n = n(),
    mass = mean(mass, na.rm = TRUE)
  ) %>%
  filter(
    n > 1,
    mass > 50
  )
```

Here this is in base R

Tapply can group data, but it only allows one function,  we could call tapply several times
and merge the data



```{R}
x=tapply(df_starwars$species,df_starwars$species,"length")
y=tapply(df_starwars$mass,df_starwars$species,"mean")
data.frame("n"=x,"mean mass"=y)
```

# sqldf

This package allows you to use SQL queries on an R dataframe.

See 

https://cran.r-project.org/web/packages/sqldf/sqldf.pdf

If you are really good at SQL,  this allows you to use your SQL chops in r

The installation was a bit cranky,  use this sequence of installs

install.packages("gsubfn")
install.packages("RSQLite")
install.packages("sqldf",type="binary")

```{R}
require('sqldf')
```
With this in place, we can run SQL commands on an R dataframe

sqldf is not tolerant of NAs,   if the df has an NA,  it will crash

the df_starwars used above has NAs,  we will use mtcars as an example

```{R}
sqldf("Select * from mtcars ")
```
```{r}
sqldf("Select *  from mtcars where mpg>=30")
```