---
title: "Set Operators in R"
author: "HDS, Mike Kozlowski"
date: "6/28/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Set Operations and Formulas in R


## Set Operations in R

HDS,  no coverage in van der Loo and de Jonge

R has a number of operations on sets, where the sets are simply vector
If x and y are vectors then we have set operations

union(x,y)
intersect(x,y)  - intersection of x and y

setdiff(x,y)
setequal(x,y)

We can also check to see if values (v) are in a given set x
is.element(v,x)   or v %in% x

```{r}
x=c(1,2,3,4,5)
y=seq(from=2,to=10, by=2)
print(x)
print(y)
```

```{r}
union(x,y)
```


```{r}
intersect(x,y)
```
Here is set difference- It returns the items in the first argument that are not in the second argument- note that it is asymmetric
```{r}
setdiff(x,y)
setdiff(y,x)
```
Compare two entire sets

```{r}
setequal(x,y)
```
We can check to see if an element (6 in this case) is an element of a set x or y

```{r}
is.element(6,x)
is.element(6,y)
6 %in% y

```

###A Few other set-like utility functions

Extracting the unique elements from a vector (or rows from a matrix or data frame)

```{r}
z=c(1,2,1,1,2,3,1,2,2,2)
unique(z)
```
We can find out the indices of the duplicated entries- the first version is not marked as duplicate
```{r}
duplicated(z)
```

Using Set Operations on the Buffalo Tax Assessment set

Get the data first

```{r}
mydata=read.csv("C:\\Users\\Mike\\Documents\\DAT511\\2-1 class\\2020-2021_Assessment_Roll.csv",header=TRUE)
summary(mydata)
```

What police districts are there in this data set?
```{r}
unique(mydata$POLICE.DISTRICT)

```

How many items are on the role

```{r}
dim(mydata)
```
How many different people own property in Buffalo

Owner1 is currently a factor, to use unique on it we must convert it to a character string
```{r}
temp_owner=as.character(mydata$OWNER1)
length(unique(temp_owner))
```
So 66,121 distinct owners of 94105 properties in Buffalo

What neighborhoods are in Buffalo?

```{r}
unique(as.character(mydata$NEIGHBORHOOD))
```

Does anyone who owns property in Hamlin Park also own property in the First Ward?

first get the lists of owners in each neighborhood

convert each to a character string form

```{r}
owners_HP=mydata$OWNER1[mydata$NEIGHBORHOOD=="Hamlin Park" ]
owners_HP=as.character(owners_HP)

owners_FW=mydata$OWNER1[mydata$NEIGHBORHOOD=="First Ward" ]
owners_FW=as.character(owners_FW)

owners_both=intersect(owners_HP, owners_FW)

owners_both

```


##writing formulas

Suppose I think the property values should depend on the living area, and I want to model that

```{r}

property_model=lm(TOTAL.VALUE~TOTAL.LIVING.AREA, data=mydata)
summary(property_model)
```
Hmm, interesting, but the R-squared is really low, .001074

Let's try a more complex model

```{r}
mydata$OVERALL.CONDITION=as.factor(mydata$OVERALL.CONDITION)
property_model=glm(TOTAL.VALUE~TOTAL.LIVING.AREA+FRONT*DEPTH+COUNCIL.DISTRICT+X..OF.KITCHENS+OVERALL.CONDITION, data=mydata)
summary(property_model)
anova(property_model,test="F")
```
