---
title: "R Notebook"
output: html_notebook
author: "Mike Kozlowski"
---

##R Loops and Control Code

R has the standard if operation to make decisions,

if the test condition is true, then the block of code delimited by the curly brackets {} is executed

```{r}

x=5

if(x<6)
{
  print("X was less than 6")
  cat("The value of X squared is :",x^2, sep=" ")
}

```

We can add an else statement, which runs if the if statement is not true

```{r}
x=7

if(x<6)
{
  print("X was less than 6")
  cat("The value of X squared is :",x^2, sep=" ")
}else
{
  print("X was greater than 6")
  cat("The value of X cubed is :",x^2, sep=" ")
}
```
Note that R is fussy about the positioning of else, it must be adjacent to the bracking } closing the if

There is an else if option as well

```{r}
x=12

if(x<6)
{
  print("X was less than 6")
  cat("The value of X squared is :",x^2, sep=" ")
}else if (x<10){
  print("X was greater than 6 but less than 10")
  cat("The value of X cubed is :",x^3, sep=" ")
}else
{
  print("X was greater than 10")
  cat("The value of square root X is :",x^0.5, sep=" ")
}

```

##For loops 

-are always iterations through a list of values

cat()- a print and concatenate function

note the use of the command code "\n",  which is a newline character

```{r}
x=c(1,2,5,11,13)
for (i in x)
{
  cat("I is:",i, "I squared is:",i^2,"\n", sep=" ")
}

```

We can compute values such as N-factorial with a loop,   this is a classic programming task

```{r}
N=5
Nfact=1
for (x in 1:N)
{
  Nfact=Nfact*x
}
print(Nfact)
```

But in R, we can often carry out tasks such as this one using built-in functions

```{r}
N=5
Nfact=prod(seq(1,N))
Nfact
```
There is also a while loop available,  it continues while the condition is true

What power of 2 is greater than 3716?

```{r}
N=0
v=2^N
target=3716
while(v<target)
{
  N=N+1
  v=2^N
}
Nlast=N-1
vlast=2^Nlast
cat("last valid N was ",Nlast," 2^N is ",vlast,"\n",sep=" ")

```

One interesting use of loops is with strings,

Here I will get the directory listing  of the current direction, and print it out

```{r}
# show the current working directory, using get working directory
cat(getwd(),"\n")
# get the directory output and store it in temp
temp=dir()
# print each file name in the directory
for (fname in temp)
{
  cat(fname,"\n")
}

```
