---
title: "Week 3 HW 2"
author: "Mike Kozlowski"
date: "2023-02-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Using Tax Assessment Roles:
```{r}
mydata = read.csv("C:\\Users\\Mike\\Documents\\DAT511\\2-1 class\\2020-2021_Assessment_Roll.csv", header = TRUE)
```

a) How many Council Districts are there?   What is the mean, median and standard deviation of property values for each council district?  What is the mean condition by Council District?
```{r}
length(unique(mydata$COUNCIL.DISTRICT))
aggregate(mydata$TOTAL.VALUE, list(mydata$COUNCIL.DISTRICT), FUN=mean)
aggregate(mydata$TOTAL.VALUE, list(mydata$COUNCIL.DISTRICT), FUN=median)
aggregate(mydata$TOTAL.VALUE, list(mydata$COUNCIL.DISTRICT), FUN=sd)
aggregate(mydata$OVERALL.CONDITION, list(mydata$COUNCIL.DISTRICT), FUN=mean)
```

b) What year appears to be the year the oldest buildings still on the tax rolls were built?  There are some properties with a year of 0 listed,  which is some form of error, remove these. Which of these has the highest assessment?  Who owns it?
```{r}
oldest = (mydata[mydata$YEAR.BUILT > 1,])
min(oldest$YEAR.BUILT, na.rm = TRUE)

mydata[mydata$YEAR.BUILT == 1780, ]
```