---
title: "Midterm Exam"
author: "Mike Kozlowski"
date: "2023-04-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require("ggplot2")
require("reshape2")
require("forcats")
require("data.table")
```

Load the data file "IBM_HR-Employee-Attrition.csv"

Generate the summary and determine if there are NAs in the data set.

Create a bar plot that shows the continuous variables DailyRate, MonthlyIncome, MonthlyRate on a single plot,  the y-axis should be log transformed.

Hint: There is an example of this is the powerpoints, early in the semester, I think, or a google search on "r plot multiple boxplot in one graph.  You will also need to do some digging in ggplot to figure out how to log transform the y axis

```{r}
ibm_infile="F:\\Chrome Downloads\\IBM_HR-Employee-Attrition.csv"
ibmdata=read.csv(ibm_infile,stringsAsFactors=TRUE)
summary(ibmdata)
sum(is.na(ibmdata))

#There are no values set as NA in the data set based on the results of the summary and the sum of is.na being 0.
```
```{r}
ibmdata_long <- melt(ibmdata, measure.vars = c("DailyRate", "MonthlyIncome", "MonthlyRate"))
ggplot(ibmdata_long, aes(x = variable, y = value)) + geom_boxplot() + scale_y_log10()
```

Load the mtcars dataset (one of the built-in data sets).

  a) Create a table that shows the mean hp of cars within each category of cylinders
  b) Create a second table that shows for each combination of category of cylinders and type of transmission, the mean quarter-mile time.

Hint: there is a function called fct_cross in tidyverse that may help here with part b- do look it up!

```{r}
data(mtcars)

sapply(split(mtcars$hp, mtcars$cyl), mean)
aggregate(qsec~am*cyl, data=mtcars, mean)

#I was unable to figure out how to get fct_cross working, so I had to take a different approach to building the second table.
```

Create an S3 structure that will hold the following information about a fish and chips shop:
  -the name
  -the owner
  -A list of the number of fish sold per month for the last 12 months
  -The pounds of potatoes used per month for the last 12 months
  -The income per month

Create a member function called plot(x) that will plot a graph of fish or potatoes over time (the last 12 months) depending on whether x is "fish" or "potatoes"

Pick values at random for the fish and potatoes entry or use rnorm() to fill them in

```{r}
mys3 <- list(name = "Long John Silver's", owner="Bob", fishMonthly=300, potatoesMonthly=200, incomeMonthly=10000)
class(mys3)<-"fishNChips"

plot.fishNChips <- function(x)
{
  if(x == "fish")
  {
    plot(x=1:nrow(mys3), y=mys3$fishMonthly)
  } else if (x == "potatoes") {
    plot(x=1:nrow(mys3), y=mys3$potatoeshMonthly)
  }
}
#plot.fishNChips("fish")

#I'm not really sure how to proceed here in getting this to function
```

Using the built-in data set "Tooth Growth", produce a graph or table that shows how the growth of pig's teeth (len) is influenced by the dose of a drug (dose) and the way the drug was delivered (sup).   Produce a single plot or table that clearly shows the impact of these two factors

```{r}
data("ToothGrowth")
ggplot(ToothGrowth, aes(x = supp, y = len)) + geom_boxplot(aes(fill = supp))+facet_grid(. ~ dose)
```