---
title: "Data Validation,  6.1"
author: "HDS"
date: "June 14, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

We are going to look at some varied data set collected by Rafael Irizarry and packaged as dslabs package

https://simplystatistics.org/2018/01/22/the-dslabs-package-provides-datasets-for-teaching-data-science/

```{r}
require("dslabs")
```
We'll start with set of student reported heights, which are supposed to be heights in inches.  This is self-reported data, with no quality control on it, so it is a mess.

This is an exercise in extracting useful information from poor quality data

This is a "craft" style attempt to fix a data set with a lot of errors, caused by people reporting height in the wrong units

We are using "domain based rules" based on what we know about human heights,  these really amount to common sense


```{r}
data("reported_heights")
str(reported_heights)
height1=tolower(reported_heights$height)
height1
```

Wow, lots of problems

-some heights look like valid values in inches
-others are in cm, or feet, or feet+inches
-some in meters
-some numbers written out as words
-some values that simply make no sense


Remove all the values written as words, replace them with numbers, this is pretty easy, we will use gsub

```{r}
height1=gsub(c("one"),c("1"),x=height1)
height1=gsub(c("two"),c("2"),x=height1)
height1=gsub(c("three"),c("3"),x=height1)
height1=gsub(c("four"),c("4"),x=height1)
height1=gsub(c("five"),c("5"),x=height1)
height1=gsub(c("six"),c("6"),x=height1)
height1=gsub(c("seven"),c("7"),x=height1)
height1=gsub(c("eight"),c("8"),x=height1)
height1=gsub(c("nine"),c("9"),x=height1)
height1=gsub(c("ten"),c("10"),x=height1)
height1=gsub(c("eleven"),c("11"),x=height1)
height1
```

What to do with the remaining values, to get height in inches?

We can state some rules about what we expect the error to be, and how to fix it

# 1.)  A single digit, followed by any non digit, then more digits must be the height in feet, plus some inches, since nobody has a height over 10 feet.

5'10, or 5 foot 10 inches or 5,10  would be interpreted as 5*12+10=70 inches

We want to find a pattern like digit, not-dight, one or more digits, the first digit is the height in feet, the second is the inches

#2.) A 1 or 2 followed by a decimal is probably height in meters

1.7, 2.15 etc

Look for a pattern of a 1 or 2, followed by a period, then 1 or 2 digits

#3.) A 3,4,5,6,or7 with a decimal is probably a decimal height in feet

5.6, 6.1 etc

Look for a pattern of a 3,4,5,6,7 followed by a period, then 1 or 2 digits

#3.)  A number in the range 100 to 250 or so is probably a height in cm



If the value is not in that pattern, we can simply strip off everything except digits and a period

The code below uses a whole series of tests using regular expresssions to look for these patterns and fix them by converting them to inches.  This was an ugly bit of work to assemble and test,  it's here to show what is possible.

Notice- we have made assumptions here about the nature of the mistake and figured out how to fix the obvious errors by converting

```{r}
height2=rep(0,length(height1))
for(i in 1:length(height1))
{
  cat("In",i,height1[i],"\n")
                             #the cat statement above was used during debugging
                              # it has been "remmed out"
  feet=0
  inches=0
  fraction=0
  cm=0
  dposition=0
  
  if(!grepl("[1-9]",height1[i]))   # if there are no numeric values present set to NA
  {
    height2[i]=NA
  }else
  if(grepl("^[1-2][0-9]{2}",height1[i]))    # three digits in a row, with a lead 1 or 2, probably                                                                                 height in cm
  {    
       # replace ,by .
       
       dposition=gregexpr("[0-9]",height1[i])  # how many digits do we have?  3 or less is height in cm
       if(length(dposition[[1]])<4)
       {   
         cm=substr(height1[i],start=dposition[[1]][1], stop=tail(dposition[[1]],1))
         height2[i]=as.numeric(cm)/2.54
       }else
       { height2[i]=NA}    # more than 3 digits in value, who knows what this means
  }else
    if(grepl("^[1-2].[0-9]{1,2}",height1[i]))   # this should be a height in meters
    {
      height1[i]=sub(',','.',height1[i])
      height2[i]=as.numeric(height1[i])*100.0/2.54
    }else
    if(grepl("^[3-7]{1}\\.[0-9]{1,3}",height1[i]))  # single digit height- feet plus fraction
      {
         height2[i]=as.numeric(height1[i])*12
      }else
  if(grepl("[1-9]{1}[^0-9]+[0-9]+",height1[i]))     # this is digit,not digit,digit
     {
       dpositions=gregexpr("[1-9]",height1[i])
       feet=substr(height1[i],start=dpositions[[1]][1], stop=dpositions[[1]][1])
       inches=substr(height1[i],start=dpositions[[1]][2], stop=tail(dpositions[[1]],1))
       height2[i]=12*as.numeric(feet)+as.numeric(inches)
     }else
    if(grepl("[1-9]{1}",height1[i])&&(!grepl("[0-9]{2}",height1[i])))   # single digit, must be height in feet
    {
       dpositions=gregexpr("[1-9]",height1[i])
       feet=substr(height1[i],start=dpositions[[1]][1], stop=dpositions[[1]][1])
       height2[i]=12*as.numeric(feet)+as.numeric(inches)
    }else
  {
    dposition=gregexpr("[0-9]",height1[i])    # first two values are height ininches,remainder are decimal place
    inches=substr(height1[i],start=dposition[[1]][1],stop=dposition[[1]][2])
    rem=dposition[[1]][-(1:2)]
    if(length(rem)>0)
    {  
      fraction=substr(height1[i],start=head(rem,1),stop=tail(rem,1))
    }else
    {
      fraction=0
    }
    height2[i]=as.numeric(paste(inches,'.',fraction,sep=''))
    cat("inches: ",i," :",height1[i],"  : ", height2[i],"\n")
  }
  #cat("out",i,height2[i],feet,inches,fraction,"\n")
  #if(!is.na(height2[i]))
  #{  
  #  if((height2[i]>80))
  #  {
  #    cat(i," :",height1[i],"  : ", height2[i],"\n")
  #  }
  #}  
  }

```

Let's see which input entries were problematic

is.na(height2) will identify the values in height2 which are set to na

We can use this set of value as an index into reported_heights$height, to see which rows are a problem 

```{r}
which(is.na(height2))
reported_heights$height[is.na(height2)]
```

Okay, I'm happy with those results, there is no way to interpret those as meaningful heights

How about checking our values a bit
we can try using a box plot and a histogram
We have a large outlier, so I used a log axis transform

```{r}
hist(height2[!is.na(height2)])
boxplot(height2[!is.na(height2)])
boxplot(height2[!is.na(height2)],log="y")
```

We have some problems, with some heights less than 24 inches and some above 96 or so
```{r}
too_short=which(height2<24)
reported_heights$height[too_short]
height2[too_short]
```

The reported heights under 24 all look pretty nonsensical

What about heights above 96 (8 feet)
```{r}
too_tall=which(height2>96)
reported_heights$height[too_tall]
height2[too_tall]
```
Remove the meaningless values

```{r}
na_list=which(is.na(height2))
bad_vals=c(na_list,too_tall,too_short)
height2[bad_vals]=NA
```

```{r}
hist(height2[!is.na(height2)],breaks=30)
boxplot(height2[!is.na(height2)])
boxplot(height2[!is.na(height2)],log="y")
```
```{r}
mod_short=which(height2<48)
reported_heights$height[mod_short]
height2[mod_short]
```
