---
title: "Validation_6_0"
author: "HDS"
date: "June 15, 2018"
output: html_document
---
updated 4/18/2023


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is a set of examples on Data Validation, using van der Loo and de Jonge, chapter 6 on validation

The chapter focuses on a sophisticated tool created by van der Loo and de Jonge for 
data validation and checking.  It is similar in intent to some commercial tools used for data validation in
large institutions.

We will use the example file "retailers" that came with the validate package as an example

see

https://cran.r-project.org/web/packages/validate/validate.pdf


We will immediately use str() to see what the data.frame contains

```{r}
require("validate")
data(retailers)
str(retailers)
```

Looking up the dataset in the manual (see URL above)

incl.prob-probability of inclusion in the sample

Turnover- this is revenue due to sales

vat-The value of turnover reported for the European Value Added Tax

size- three categories of size,   sc0 means unknown


One of the cardinal rules of working with any data set, is always do some simple plots and look at the data

```{r}
boxplot(retailers)
```

Okay, this boxplot isn't very helpful, as the ranges on the different variables are so different.

We need individual box plot for each variable

We'll do this using ggplot2 to generate the plots, and cowplot to arrange them on a page

we have a factor size that describes the size of each store in this chain of european grocery stores

```{r}
require(ggplot2)
require(cowplot)

staffPlot=ggplot(retailers,aes(size,staff))+geom_boxplot();
incl.probPlot=ggplot(retailers,aes(size,incl.prob))+geom_boxplot();
turnoverPlot=ggplot(retailers,aes(size,turnover))+geom_boxplot();
other.revPlot=ggplot(retailers,aes(size,other.rev))+geom_boxplot();
total.revPlot=ggplot(retailers,aes(size,total.rev))+geom_boxplot();
staff.costsPlot=ggplot(retailers,aes(size,staff.costs))+geom_boxplot();
total.costsPlot=ggplot(retailers,aes(size,total.costs))+geom_boxplot();
profitPlot=ggplot(retailers,aes(size,profit))+geom_boxplot();
vatPlot=ggplot(retailers,aes(size,vat))+geom_boxplot();
plot_grid(staffPlot,incl.probPlot,turnoverPlot,other.revPlot,total.revPlot,staff.costsPlot,total.costsPlot,profitPlot,vatPlot)
```
Based on these results, the data looks pretty dirty  Pretty much every variable has some pretty extreme outliers

Outliers may be rare events or they may be data errors of various types- it's always a bit of a problem to decide which they really are, and what to do about them, the validate tool is one approach

With the wide range in these data sets, we can't seem about the extent of variation within each set, what we need here is a log axis on the y-scale to make it easier to see the smaller vartions- this should be easy to do

Under ggplot2's grammar of graphics, we just add a scale modifier to each graph- cut and paste!

```{r}

staffPlot=ggplot(retailers,aes(size,staff))+geom_boxplot()+scale_y_continuous(trans="log10");
incl.probPlot=ggplot(retailers,aes(size,incl.prob))+geom_boxplot()+scale_y_continuous(trans="log10");
turnoverPlot=ggplot(retailers,aes(size,turnover))+geom_boxplot()+scale_y_continuous(trans="log10");
other.revPlot=ggplot(retailers,aes(size,other.rev))+geom_boxplot()+scale_y_continuous(trans="log10");
total.revPlot=ggplot(retailers,aes(size,total.rev))+geom_boxplot()+scale_y_continuous(trans="log10");
staff.costsPlot=ggplot(retailers,aes(size,staff.costs))+geom_boxplot()+scale_y_continuous(trans="log10");
total.costsPlot=ggplot(retailers,aes(size,total.costs))+geom_boxplot()+scale_y_continuous(trans="log10");
profitPlot=ggplot(retailers,aes(size,profit))+geom_boxplot()+scale_y_continuous(trans="log10");
vatPlot=ggplot(retailers,aes(size,vat))+geom_boxplot()+scale_y_continuous(trans="log10");
plot_grid(staffPlot,incl.probPlot,turnoverPlot,other.revPlot,total.revPlot,staff.costsPlot,total.costsPlot,profitPlot,vatPlot)
```
That's easier to see

It really looks like there is no variation in incl.prob within each size class- so whatever this variable incl.prob is, it doesn't see to mean much beyond size class

Let's look at just that plot
```{r}
plot(incl.probPlot)
```
# we can create all the biplots, just to get a quick look

```{R,fig.width=12,fig.height=12}
plot(retailers)
```



Yeah, no signs of any variation in incl.prob with size class, so we won't try to do much more with incl.prob

#PCA- Principle Component Analysis

This is one of my favorite exploratory data analysis (EDA) tools- it looks for axes that display the greatest amount of variation in the data

Outlier show up as large deviations from the general patterns, so they are likely to be points of concern for us

```{r}
mypca=prcomp(~staff+turnover+other.rev+total.rev+staff.costs+profit+vat,data=retailers,scale=TRUE,na.action=na.omit)

scores = as.data.frame(mypca$x)
ggplot(data = scores, aes(x = PC1, y = PC2, label = rownames(scores))) +
  geom_hline(yintercept = 0, colour = "gray65") +
  geom_vline(xintercept = 0, colour = "gray65") +
  geom_text(colour = "tomato", alpha = 0.8, size = 4) +
  ggtitle("PCA plot of retailer data set, NA omit")

dim(scores)
```

So on this plot, something is wrong with store # 36, it is wildly differnent from the other data points

The dim(scores) call also indicates it is 17 x 7, so we have a lot of stores with one or more na values, since we omitted the nas in the PCA plot

lets look at the NAs,  we will create a function to sum the Na values over a column
```{r}
f<-function(x){sum(is.na(x))}
apply(retailers[,3:10],2,FUN=f)
```

Other.rev is really the issue here, lots of NAs there

Let's omit other.rev and and try the pca again

```{r}
mypca2=prcomp(~staff+turnover+total.rev+staff.costs+profit+vat,data=retailers,scale=TRUE,na.action=na.omit)

scores2 = as.data.frame(mypca2$x)
ggplot(data = scores2, aes(x = PC1, y = PC2, label = rownames(scores2))) +
  geom_hline(yintercept = 0, colour = "gray65") +
  geom_vline(xintercept = 0, colour = "gray65") +
  geom_text(colour = "tomato", alpha = 0.8, size = 4) +
  ggtitle("PCA plot of retailer data set, other.revs omitted")

dim(scores2)
```
That's better, we now have 36 entries out of 60 included in the PCA- still not great

Why do 19,36 look like outliers?

```{R}
retailers[c(19,36),]
```

Okay,  19 has the largest profit by far

36 has a really high other.rev,  and a huge staff cost


#Data Summary
```{R}
summary(retailers)
```





#Automated Error Detection

Let's look at the validate package a bit

It works on a set of rules for the data, that is to say pre-determined statements we can make about the relationship of the variables within the data.

Based on van der Loo and de Jonge's comments,

    turnover>=0

    staff.costs/staff is less than 50 (50,000 fl per employee)

The check_that() function can look at these rules

```{r}
cf<-check_that(retailers,turnover>0,staff.costs/staff<50)
summary(cf)
```

To see which lines in the data frame failed the validation check, we can look at values(cf)

```{R}
values(cf)
```


Let's expand this, to check to see if there are staff, that the staff.cost is no zero, and that the operation as a whole generates a profit, note that the conditions have now been named

Note that we have a conditional statement in here (if there are staff then staff.costs>0)
and a summary statement (mean profits >0)

the conditional statement involves two variables,   staff and staff.costs, so if this condition is not met, we cannot be sure which is in error,  this is a u (item) multiple comparison

The summary statement involves all the data, it is about the entire set, not a particular line

```{r}
cf<-check_that(retailers,
               to=turnover>0,
               sc=staff.costs/staff<50,
               cd=if(staff>0) staff.costs>0,
               mn=mean(profit,na.rm=TRUE)>0)
summary(cf)

```
The variable cf is called a confrontation, and it contains information about the testing procedure

We can also plot it


```{r}
barplot(cf[-4],main="Retailers")
```
