---
title: "Validation_More_6_5_7"
author: "HDS"
date: "August 8, 2018"
output: html_document
---
checked 4/18/2023

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## More Validation Examples

following van der Loo and de Jonge,  section 6.5.7 and onwards

Specific examples of how to test specific aspects of data sets

###6.5.7 Variable types and metadata

Are data of the correct class? Check to make sure the data types are all correct

We can use class() and the standard is.{something} functions

Note: I'm using check_that() here for demonstration purposes, since these are single line rules 

```{r}
require("validate")
data("retailers")

check_that(retailers,is.integer(staff))
check_that(retailers,class(total.rev)=="int")
```

Check the number of rows in a data set

Looking at the iris data set here

```{r}
check_that(iris,nrow(.)>=150)
```

Check the percentage of missing data
```{r}
check_that(retailers, sum(is.na(.))/prod(dim(.))<0.2)
```
###Values in lists of allowed values

these are often string comparisons

gender %in% c("male","female")

tolower(gender)%in% c("male","female")

or use fuzzier matching- refer to chapter 5 on matching criteria

You can create lists within the validator object to store lists of allowed values,  using the := operator, which is part of the validate DSL
```{r}
require('magrittr')
d<-data.frame(gender=c("male","female","female","male","unknown"))
v<-validator(
  gendercodes:=c("male","female","f","m"),
  GV=gender %in% gendercodes
)
confront(d,v) %>%summary()
```


###Cross Record Dependencies 6.5.10

May change when the data set is modified- this may make it difficult to localize the error

```{r}
v<-validator(
  cvcheck=cov(height,weight)>0
)
cf=confront(women,v)
summary(cf)
```

```{r}
v<-validator(
  height<(median(height)+1.5*IQR(height))
)
cf=confront(women,v)
summary(cf)
```

But if this fails, we would want to know why.  Was a height too high?  was there a problem with the estimated median or the inter-quartile range (IQR) from the data?    In other words, is the problem one data point, or the set as a whole?



###Cross data set validation

Using data from two or more sets

We can pass a list or a data frame in as a reference, and then use values or functions of values in the list as a validation rule

Only one list may be sent, but one could place many dataframes within a list

We can compare conditions in one set to parameters derived from another set

```{r}
males=data.frame( height=c(70,68,71,67),weight=c(180, 190, 145, 192))
v<-validator(
  women_mean_h:=mean(height),
  upbound=women_mean_h<mean(ref$height),
  lowbound=women_mean_h>0.85*mean(ref$height)
)
cf=confront(dat=women,x=v,ref=males)
summary(cf)
```

###Macros,  Variable groups and Keys in Validate  6.5.13

These are features of the DSL meant to make it easier to create rules and re-use code

Macros- using the := operator within the validator construction

x:=6,   so x is a stored vlaue
y:=mean(x,na.rm=TRUE)   is a stored macro, so it can be re-used in the code, helpful if it is used over and over

z<y  is read as z<mean(x,na.rm=TRUE)

Variable groups

Allows you to apply the same rule to all variables in the group

var_group(a,b)>0   means two rules,   a>0 and b>0

###Analyzing Output: validation Objects 6.5.14

The output of a confrontation can be large,  we would like to aggregate results to make them easier to follow

Here's an example from the retailers set

There is no good key in this data set right now, we will add an identifier, we then send this to confront as the key
```{r}
retailers$id=paste0("r-",1:nrow(retailers))
v<-validator(
  OTREV=other.rev>0,
  TOver=turnover>0,
  TRev=total.rev>0,
  StfCost=staff.costs>0,
  TC=total.costs>0,
  RevBal=turnover+other.rev==total.rev
)
cf<-confront(retailers,v,key="id")
summary(cf)
```

```{r}
aggregate(cf)
```
We can see where the values occurred

```{R}
values(cf)
```



We can aggretate by record, to see the failures by record line

This is often what we want to see, which records indicate failures, although we know for some rules, errors are hard to assign to a particular record

```{r}
t=aggregate(cf,by="record")
head(t,n=5)
```

We can sort by records as well

```{r}
t=sort(cf,by='record')
head(t,n=5)
```

Once we have t, we could slice out all the records with less than 6 passes, ie all the records that had at least 1 failure or NA

```{r}
t[t$npass<6,]
```


